|

### 2024-08-22 reproducibility/manuscript/manuscript.qmd
docs(manuscript): update methods
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@

Given observations $\tilde{X}_{cg} = \left( u_{cg}^{obs}, s_{cg}^{obs} \right)$,
we would like to compute the posterior distribution over the random variables
\begin{align}
\theta &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
\end{align}
but exact Bayesian inference is intractable in this model. We use Pyro to
automatically integrate out the local discrete latent variables $k$, which is
defined as the cell and gene transcriptional state (see above), and approximate
the posterior over the remaining latent variables using variational inference
[@Bingham2018-id;@Kucukelbir2016-bk], which converts intractable integrals into
optimization problems that can be solved with off-the-shelf tools. In
variational inference, we maximize a tractable bound (the evidence lower bound
(ELBO)) on the Kullback-Leibler (KL) divergence
$$\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} p(\theta, \psi \vert \tilde{X}_{cg}) \right)$$
between a parametric family of tractable probability distributions q and the
intractable true posterior
\begin{align}
\varphi^* &= \operatorname{argmax}_{\varphi} \text{ELBO} \nonumber \\
&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
&\qquad\qquad\qquad\qquad -\log \left(q_{\varphi}(\theta, \psi)\right) \bigg] \bigg\}
\end{align}
We approximate our model’s posterior distribution $p( \theta, \psi \vert
\tilde{X})$ with a tractable family of probability distributions
$q_{\psi}(\theta, \phi) = q_{\phi_1}(\theta) q_{\phi_2}(\psi)$, where
$q_{\phi_1}(\theta)$ is a product of independent univariate Gaussian
distributions with learnable location and scale parameters and
$q_{\phi_2}(\psi)$ is a multivariate Gaussian distribution with a learnable
location parameter and a learnable low-rank covariance matrix. We solve the
resulting stochastic optimization problem using a version of Automatic
Differentiation Variational Inference (ADVI) [@Kucukelbir2016-bk]: we obtain
gradient estimates by differentiating a Monte Carlo estimate of the ELBO and
update the variational parameters  using stochastic gradient ascent.  Our
implementation and experiments use generic variational families and Monte Carlo
ELBO and gradient estimators provided by Pyro [@Bingham2018-id], an open source
library for probabilistic machine learning, and a version built into Pyro of the
adaptive stochastic gradient ascent algorithm Adam augmented with gradient
clipping to enhance numerical stability during training.

@@

We implemented Cytotrace in Python based on the R version [@Gulati2020-xq] for
benchmarking Pyro&thinsp;-Velocity and scVelo’s shared (latent) time prediction in the
Python environment. This implementation is available in the
[Pyro&thinsp;-Velocity GitHub repository](https://github.com/pinellolab/pyrovelocity/blob/main/src/pyrovelocity/analysis/cytotrace.py).

@@

Pyro&thinsp;-Velocity provides two model formulations as described above.
Users can compare their predictive performance to decide which of the two models
to use for a given dataset. Namely, the mean absolute errors of posterior
samples prediction of spliced and unspliced raw read counts $x_{icg}$ from its
observation $\tilde{X}_{cg}$ per gene and cell combinations
\begin{align*}
\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} |x_{icg} - \tilde{X}_{cg}|,
\end{align*}
where $i$ is the posterior sample index, $N_i$ is the posterior sample number,
$N_c$ is the cell number, $N_g$ is the gene number. Then it is possible to
evaluate the cell fate inference performance of the models based on two metrics:

1. the consistency between the velocity vector field and the *clonal progression
vector field* using grid-based cosine similarity; 2. Spearman’s correlation of
velocity-based shared (latent) time with Cospar fate potency and the silver
standard Cytotrace score. Both metrics are implemented in scipy.

@@

L.P. has financial interests in Edilytics, Inc., Excelsior Genomics, and SeQure
Dx, Inc. L.P.’s interests were reviewed and are managed by Massachusetts General
Hospital and Partners HealthCare in accordance with their conflict of interest
policies.

@@

Extension of @fig-lineage-tracing, Pyro&thinsp;-Velocity (Model 1 for uni-fate
data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on
scRNA-seq datasets with lineage barcoding information (LARRY dataset). **a**, SPRING
embedding rendering of Cytotrace scores, the title reports the Spearman's
correlation between the Cytotrace and Cospar fate potency scores, **b**,
Pyro&thinsp;-Velocity additional single cell level uncertainty diagnostic plots
based on the multi-fate LARRY dataset, from left to right: 1. Standard deviation
of posterior samples of shared time per cell, 2. Rayleigh test of posterior
samples of vector field per cell (the title reports the false discovery rate
after p-value correction for multiple tests across cells), 3. Transcriptional
state uncertainty of predicted raw read count per cell. The contour lines show
the top 10% of the most uncertain cells ([Methods](#sec-methods)), **c**, Pairwise Spearman's
correlation between Cytotrace/Cospar fate potency scores and velocity methods
for the four different benchmarking datasets selected based on the lineage
barcoding information from LARRY data, **d**, SPRING embedding rendering of
posterior mean of shared time across 30 posterior samples, the title shows the
Spearman's correlation of mean shared time with Cytotrace. **e**, Selection of the
top positively correlated genes with the cell shared time based on the posterior
mean of spliced expression. **f**, Phase portraits, rainbow plots, and UMAP
rendering with splicing gene expression level for the selected genes in **e**.
:::

>>>>

We use these solutions to formulate an end-to-end probabilistic generative model 
that relates prior distributions on kinetic parameters to a distribution on pairs 
of observed unspliced and spliced read count matrices

\begin{align}
\alpha^{(0)}{ }_g &\sim \operatorname{LogNormal}(0,1), \\
\beta_g &\sim \operatorname{LogNormal}(0,1), \\
\gamma_g &\sim \operatorname{LogNormal}(0,1), \\
&\hskip -18pt \Delta \text { switching }_g \sim \operatorname{LogNormal}(0,1), \\
t_0^{\left(k_{c g}\right)} &= \left\{
  \begin{array}{l}
    t_0^{(0)}{ }_g \sim \operatorname{Normal}(0,1), k_{c g}=0 \\
    t_0^{(1)}{ }_g=t_0^{(0)}{ }_g+\Delta \text { switching }_g, \\
    \quad k_{c g}=1
  \end{array}\right. \\
t_c &\sim \operatorname{LogNormal}(0,1), \\
k_{c g} &\sim \text{Bernoulli} \left( \text{logits}= t_c-t_0^{(1)} \right), \\
\tau^{\left(k_{c g}\right)} 
  &= \operatorname{softplus}\left(t_c-t_0^{\left(k_{c g}\right)}{ }_g\right), \\
u_{c g} 
 &= \text { Measurement }_u \left( u\left(\tau^{\left(k_{c g}\right)}\right) ; 
                                   u_{c g}^{obs}\right), \\
s_{c g} 
  &= \text { Measurement }_s \left( s\left(\tau^{(k_{c g})}\right) ; 
                                    s_{c g}^{obs}\right).
\end{align}
$u\left(\tau^{\left(k_{c g}\right)}\right)$ and $s\left(\tau^{(k_{c g})}\right)$ are
are called the denoised gene expression calculated from the velocity analytic
solution input with the kinetics random variables. $u_{cg}$ and $s_{cg}$ are the spliced
and unspliced read count sampled from the Poisson models. $u_{cg}^{obs}$ and $s_{cg}^{obs}$ are
the observed spliced and unspliced read count tables. The generative process

$\text{Measurement}(\cdot)$ for observed unspliced read counts given denoised unspliced
gene transcript expression level $u\left(\tau^{(k_{cg})}\right)$ 
(and identical for observed spliced read counts) models the expected number of observed 
reads for a given gene in a given cell as the number of transcripts times the ratio of 
the cell's total reads to total transcripts
\begin{align}
u_c^{\hat{obs}} &= \sum_g u_{c g}^{obs}, \\
\hat{u}_c &= \sum_g u\left( \tau^{(k_{c g})}\right), \\
\eta_c^{(u)} &\sim \operatorname{Normal}\left(
    u_c^{\hat{obs}_c}, 
    \operatorname{std} \left(u_c^{\hat{obs}}\right)
  \right), \\
\mu_{c g}^{(u)} &= \log \left(u\left(\tau^k{ }_{c g}\right)\right)
  +\log \left(\eta_c^{(u)}\right)-\log \left(\hat{u}_c\right), \\
u_{c g}^{obs} &\sim 
  \operatorname{Poisson}\left(\lambda=\exp \left(\mu_{c g}^{(u)}\right)\right).
\end{align}

For the first Pyro-Velocity model (Model 1), we constrain the shared time to be strictly
larger than $t_{0}^{(0)}$ by introducing auxiliary random variables 
$$
\text{t\_constraint}_{cg} 
  \sim \text{Bernoulli} \left( \text{logits} = t_c - {t_{0}^{(0)}}_g \right),
$$
and setting their values to $1$, and we set the initial condition per gene to be
\begin{align}
\left( {u_{0}^{(k_{cg})}}_g , {s_{0}^{(k_{cg})}}_g \right) &= \left\{
  \begin{array}{l}
    (0,0), k_{c g}=0 \\
    \bigg( {u \left( \Delta \text { switching }_g \right)}_g,\\
           \quad {s \left( \Delta \text { switching }_g \right)}_g \bigg), \\
           \quad k_{c g}=1
  \end{array}\right.
\end{align}
For the extended Pyro&thinsp;-Velocity model (Model 2), we remove the shared
time constraint $\text{t\_constraint}_{cg}$, thus allowing a time lag per gene
that might be caused by delayed gene activation and set the initial condition
per gene as random variables that are strictly positive $\left({u_{0}^{(0)}}_g,
{s_{0}^{(0)}}_g\right)$, which allow genes having a basal expression level before gene
activation. Then, we compute the gene expression at the second switching time as
\begin{align}
({u_{0}^{(1)}}_g, {s_{0}^{(1)}}_g) &= 
  \bigg( {u \left( \Delta \text { switching }_g \right)}_g, \nonumber \\
& \qquad {s \left( \Delta \text { switching }_g \right)}_g \bigg),
\end{align}
which shares the same initial condition $\left({u_{0}^{(0)}}_g, {s_{0}^{(0)}}_g\right)$ where
\begin{align}
{u_{0}^{(0)}}_g &\sim \operatorname{LogNormal}(0,1),\\
{s_{0}^{(0)}}_g &\sim \operatorname{LogNormal}(0,1).
\end{align}

@@

Given observations $\tilde{X}_{cg} = \left( u_{cg}^{obs}, s_{cg}^{obs} \right)$,
we would like to compute the posterior distribution over the random variables
\begin{align}
\theta &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, 
          {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
\end{align}
but exact Bayesian inference is intractable in this model. We use Pyro to
automatically integrate out the local discrete latent variables $k$, which is
defined as the cell and gene transcriptional state (see above), and approximate
the posterior over the remaining latent variables using variational inference
[@Bingham2018-id;@Kucukelbir2016-bk], which converts intractable integrals into
optimization problems that can be solved with off-the-shelf tools. In
variational inference, we maximize a tractable bound (the evidence lower bound
(ELBO)) on the Kullback-Leibler (KL) divergence
$$
\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} 
  p(\theta, \psi \vert \tilde{X}_{cg}) \right)
$$
between a parametric family of tractable probability distributions q and the
intractable true posterior
\begin{align}
\varphi^* &= \operatorname{argmax}_{\varphi} \text{ELBO} \nonumber \\
&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} 
     \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
&\qquad\qquad\qquad\qquad -\log \left(q_{\varphi}(\theta, \psi)\right) \bigg] \bigg\}
\end{align}
We approximate our model's posterior distribution $p( \theta, \psi \vert
\tilde{X})$ with a tractable family of probability distributions
$q_{\psi}(\theta, \phi) = q_{\phi_1}(\theta) q_{\phi_2}(\psi)$, where
$q_{\phi_1}(\theta)$ is a product of independent univariate Gaussian
distributions with learnable location and scale parameters and
$q_{\phi_2}(\psi)$ is a multivariate Gaussian distribution with a learnable
location parameter and a learnable low-rank covariance matrix. We solve the
resulting stochastic optimization problem using a version of Automatic
Differentiation Variational Inference (ADVI) [@Kucukelbir2016-bk]: we obtain
gradient estimates by differentiating a Monte Carlo estimate of the ELBO and
update the variational parameters  using stochastic gradient ascent.  Our
implementation and experiments use generic variational families and Monte Carlo
ELBO and gradient estimators provided by Pyro [@Bingham2018-id], an open source
library for probabilistic machine learning, and a version built into Pyro of the
adaptive stochastic gradient ascent algorithm Adam augmented with gradient
clipping to enhance numerical stability during training.

@@

We implemented Cytotrace in Python based on the R version [@Gulati2020-xq] for
benchmarking Pyro&thinsp;-Velocity and scVelo's shared (latent) time prediction in the
Python environment. This implementation is available in the
[Pyro&thinsp;-Velocity GitHub repository][cytotrace].

[cytotrace]: https://github.com/pinellolab/pyrovelocity/blob/main/src/pyrovelocity/analysis/cytotrace.py

@@

Pyro&thinsp;-Velocity provides two model formulations as described above.
Users can compare their predictive performance to decide which of the two models
to use for a given dataset. Namely, the mean absolute errors of posterior
samples prediction of spliced and unspliced raw read counts $x_{icg}$ from its
observation $\tilde{X}_{cg}$ per gene and cell combinations
\begin{align*}
\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} 
  \vert x_{icg} - \tilde{X}_{cg} \vert,
\end{align*}
where $i$ is the posterior sample index, $N_i$ is the posterior sample number,
$N_c$ is the cell number, $N_g$ is the gene number. Then it is possible to
evaluate the cell fate inference performance of the models based on two metrics:

1. the consistency between the velocity vector field and the *clonal progression
vector field* using grid-based cosine similarity; 2. Spearman's correlation of
velocity-based shared (latent) time with Cospar fate potency and the silver
standard Cytotrace score. Both metrics are implemented in scipy.

@@

L.P. has financial interests in Edilytics, Inc., Excelsior Genomics, and SeQure
Dx, Inc. L.P.'s interests were reviewed and are managed by Massachusetts General
Hospital and Partners HealthCare in accordance with their conflict of interest
policies.

@@

Extension of @fig-lineage-tracing, Pyro&thinsp;-Velocity (Model 1 for uni-fate
data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on
scRNA-seq datasets with lineage barcoding information (LARRY dataset). **a**,
SPRING embedding rendering of Cytotrace scores, the title reports the Spearman's
correlation between the Cytotrace and Cospar fate potency scores, **b**,
Pyro&thinsp;-Velocity additional single cell level uncertainty diagnostic plots
based on the multi-fate LARRY dataset, from left to right: 1. Standard deviation
of posterior samples of shared time per cell, 2. Rayleigh test of posterior
samples of vector field per cell (the title reports the false discovery rate
after p-value correction for multiple tests across cells), 3. Transcriptional
state uncertainty of predicted raw read count per cell. The contour lines show
the top 10% of the most uncertain cells ([Methods](#sec-methods)), **c**,
Pairwise Spearman's correlation between Cytotrace/Cospar fate potency scores and
velocity methods for the four different benchmarking datasets selected based on
the lineage barcoding information from LARRY data, **d**, SPRING embedding
rendering of posterior mean of shared time across 30 posterior samples, the
title shows the Spearman's correlation of mean shared time with Cytotrace.
**e**, Selection of the top positively correlated genes with the cell shared
time based on the posterior mean of spliced expression. **f**, Phase portraits,
rainbow plots, and UMAP rendering with splicing gene expression level for the
genes selected in **e**.
:::
